Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-01-10 RefereeResponse/RefereeResponse_v2.tex

\documentclass[11pt]{article}

\usepackage[usenames,dvips]{pstcol}

\usepackage{color}

\usepackage{amssymb}

\usepackage{amsmath}

%

\input{symbols.tex}

%

\newcounter{question}

\newcommand{\noteq}{{\vskip0.1in\bf \large \color{red} [Q:{\color{black}\addtocounter{question}{1}\arabic{question}}]~}}

\newcommand{\noteQ}{{\bf \LARGE \color{blue} [Q: {\color{black}\addtocounter{question}{1}\arabic{question}}]~}}

\newcommand{\notea}{{\bf \large \color{red} [A]~}}

\newcommand{\noteA}{{\bf \LARGE \color{blue} [A]~}}

\newcommand{\noteqq}[1]{{\bf \large \color{red} [Q: {\color{black}#1}]~}}

\newcommand{\noteaa}[1]{{\bf \large \color{red} [A: {\color{black}#1}]~}}

%

\textwidth 6.5in

\textheight 9.4in

\topmargin -0.8in

\oddsidemargin 0.0in

\evensidemargin 0.0in

\parindent 0.0in

\def\baselinestretch{1.0}

\def \Russ {Translated from Russian. Fixed. Thanks.}

\def \F    {Fixed.}

\def \FT   {Fixed. Thanks.}

\def \D    {Done.}

\def \AP   {Basically Stylistic- Author's Prerogative.}

\begin{document}

\begin{flushright}

\today

\end{flushright}

%----------------------------------------------------------------

Dear JINST Editors,\\

\parindent 1.0in

We thank the referees for their very helpful and knowledgeable

comments. We have substantially rewritten the manuscript in response

to the comments as detailed in the list below.

There are several areas, however, in which the referees requested

substantially expanding the scope of the paper.  We envisioned this

paper as a proof of  the principle that the use of the fast timing

photodetectors presently being developed would enable the separation

of Cherenkov and scintillation light, as the reconstruction of the

direction of low energy ($\sim$1 MeV) electrons in a liquid

scintillator detector has never been shown before.  In this paper, we

want to: 1) show scintillation/Cherenkov separation is possible; 2)

determine the parameters of a large scintillator detector that are

important to this separation; and 3) demonstrate that reconstruction

techniques based on Cherenkov light would work in idealized

situations.

We are now working on two follow-up papers that are substantially

beyond the scope of the present paper, moving beyond

proof-of-principle into the laborious and time-intensive area of

reconstruction algorithms, areas for which the difficulty is known

well by some of us from our experience on KamLand. The first is a

detailed study of reconstruction in more realistic scenarios. The

second is a proposal for a $0\nu\beta\beta$ experiment  where the detailed

background reduction factors and sensitivity to new physics are

presented, SuperNEMO has a similar paper, but for scintillator

detectors nothing similar exists since a directional signal has not

been considered for them in the past.  Finally, we will be having a

turnover of young personnel so this paper provides an accessible summary

of their work.

\parindent 3.0in

    Sincerely,\\

\parindent 4.0in

		         Lindley Winslow, for the Authors

\parindent 0.0in

%\begin{verbatim}

%\end{verbatim}

\noteq [Referee 1 Intro] ``I like  very much the clean, transparent, simple

and well-motivated methodology (with one exception..). The structure

of the paper is nice''

\notea{\bf Many thanks.}\\

\noteq [Referee 1 Intro] ``However the presentation of the results is a

different thing... I am particularly concerned with the fact...,

instead of showing the resulting CosTheta distribution that would seem

like the obvious result one wants to see. .... Why not to disclose

this openly to allow the reader to evaluate the results in detail?''

\notea{\bf  This is our fault- the desired plot was (is) there

(Figure 3), but was labeled as $p_x/|\vec{p}|$ rather than the obvious

CosTheta that it is. We have made this clear on the figure and in the

text. Thanks.}\\

\noteq [Referee 1: Page 1] ``Please specify FWHM or sigma and give a

reference for both cases.''

\notea{\bf  \D }\\

DOES ANYBODY KNOW WHAT HSE IS ASKING FOR WRT THE REFERENCE?

\noteq [Referee 1: Page 1] ``...but please avoid misplacing the adjective good in

this context. ...''

\notea{\bf  \FT}\\

\noteq [Referee 1: Page 2] ``..the authors should..check and refer to

the work of NEXT and Super-Nemo collaborations:''

\notea{\bf  \D}\\

I DON'T SEE THE REFERENCE TO NEXT IN THE TOPOLOGY PARAGRAPH- DID I MISS

IT? IT SHOULD GO AFTER ``..ESPECIALLY FOR 0NUBETABETA SEARCHES'''?

\noteq [Referee 1: Page 2] ``Define PPO here, not later''

\notea{\bf \D}\\

\noteq [Referee 1: Page 3] ``Please define all magnitudes in eq. 2.1 and 2.2''

\notea{\bf \D}\\

\noteq [Referee 1: Page 3]  ``Indicating the direction of the

electrons should not harm''

\notea{\bf  \D}\\

WAS ANYTHING CHANGED IN THE FIGURE? THE CAPTION MAKES IT CLEAR, AND

THE BRIGHT WHITE AXIS ARROW IS CLEAR... IF WE DON'T CHANGE THE FIG, WE

SHOULD SAY SO AND WHY NOT (EASY)

\noteq [Referee 1: Page 3] ``None of the best limits or claims in bb0

sues an isotope that satisfies this condition..''

\notea{\bf \FT}\\

IS THIS IS FIXED? I COULDN'T FIND IT IN THE CURRENT DRAFT AROUND

THE 2.6 MEV DISCUSSION. .I'M CONFUSED BY THE COMMENT, FRANKLY. DOES

THE PRESENT TEXT NEED MORE WORK? (I.E. I NEED EDUCATING...)

\noteq [Referee 1: Page 3]

\notea{\bf \FT}\\

THIS IS FIXED, RIGHT? I COULDN'T FIND IT IN THE CURRENT DRAFT AROUND

THE 2.6 MEV DISCUSSION. .I'M CONFUSED BY THE COMMENT, FRANKLY. DOES

THE TEXT NEED WORK? (I.E. I NEED EDUCATING...)

\noteq [Referee 1: Page 4] ``could you perhaps include the

attenuation @400nm and n@400..?''

\notea{\bf \D}\\

I SEE THE ATTENUATION LENGTH, BUT NOT N@400

\noteq [Referee 1: Page 4] Mentioning the figure in the citation.

\notea{\bf ?}\\

I DIDN'T UNDERSTAND THIS COMMENT- THE PAPER MENTIONS FIG 4C..

\noteq [Referee 1: Page 4] ``Could you please include also the cutoff

wavelength previously referred to?''

\notea{\bf ?}\\

I DIDN'T UNDERSTAND THIS ONE EITHER

\noteq [Referee 1: Page 4] `` Scattering have not been included..''

\notea{\bf We have added a figure and accompanying text to show that

  the scattering only adds a late tail. Thanks.}\\

\noteq [Referee 1: Page 4] ``For which digitized values come from ??''

\notea{\bf \F}\\

\noteq [Referee 1: Page 4] ``5 MeV electrons.. It is very disturbing

to benchmark the simulation with electrons having an energy doubling

the Qbb of the most common isotopes..''

\notea{\bf We have clarified in several places

that the 5 MeV case is for neutrino-electron experiments. For dbb we

have made it explicit that we consider 1.4 and 2.1 MeV}\\

\noteq [Referee 1: Page 4] ``Please specify also the track length in

the direction of emission''

\notea{\bf \D}\\

\noteq [Referee 1: Page 5] ``Before the solutes in liquid... Please rephrase''

\notea{\bf \D}\\

\noteq [Referee 1: Page 5] ``I hope that [43] is not the final version

but a draft.

\notea{\bf It was a draft; we have changed the reference. Thanks!}\\

\noteq [Referee 1: Page 6] ``So I would be very much convinced ..if

the authors would report Fig. 3 during the electron history

(equivalently, energy or distance, including several histograms for

each case. It would be nice to see at which energy scale or length the

electron isotropizes and the Cherenkov cone smears out.''

\notea{\bf We have added studies of 1.4 MeV and 2.1 MeV electrons,

  which show that the separation and directionality based on timing

  hold as the electron energy is lowered, as shown in the newly-added

  Figures 5 and 6. We have also added the path

  lengths and total distance and time from the vertex. We believe that

  additional plots beyond the ones we have added won't add any useful

  information.}\\

 HSE'S CLEARLY STUCK ON THIS- MISSED THE POINT. HOWEVER, WE NEED TO

  ANSWER THIS DIRECTLY - IS IT POSSIBLE TO MAKE ONE PLOT OF

\noteq [Referee 1: Pages 7-8] ``This is interesting, but specially

section 6 seems a bit off-topic to me. I am not against keeping it...''

\notea{\bf We'd like to keep it. }\\

\noteq [Referee 1: Page 9] ``Please add a histogram of the reconstructed

angle for all energies...''

\notea{\bf }\\

NEEDS ATTENTION...

\noteq [Referee 1: Page 9] ``..the vertex reconstruction has to be

obtained from all photons and the direction from the Cherenkov photons..''

\notea{\bf }\\

I CAN'T WRITE AN ANSWER TO THIS- CHRISTOPHE,ANDREY, MATT, LINDLEY?

(there's lots of email that contains the answer, it seems...)

\noteq [Referee 1: Fig. 7] ``the authors should show the spread of the

cos(theta) distribution for each of the energies.

\notea{\bf \D See Figure 3.}\\

I DON'T HAVE A FIGURE 7. ???

\noteq [Referee 1: Fig. 7 ``It would be nice to show/indicate..also the

Cherenkov threshold for high wave-length (n about 1.4)

\notea{\bf We have added this to the text.}\\

I DON'T HAVE A FIGURE 7. ???

\noteq [Referee 1: Conclusions] ``Should be rewritten after the

previous objections are addressed.''

\notea{\bf \D Thanks!}\\

WE SHOULD REWRITE THE CONCLUSIONS TO BE MORE SPECIFIC WITH MORE

FACTS. AS THEY STAND THEY'RE SORT OF GENERIC- IF THAT'S ALL ONE READS

OF THE PAPER ONE'S NOT MUCH WISER ...

\noteq [Referee 1: Page 7-8] ``I see no convincing proof of it with

the information here presented. Hope the answers help clarifying the issue.''

\notea{\bf We have made the sentence more specific. Thanks.}\\

WE SHOULD BE MORE SPECIFIC- CAN GIVE NUMBERS RATHER THAN OPINIONS

HERE..

%\newpage

\noteq [Referee 2: General Comments] ``The paper introduces an

original technical development which could increase the interest of

large scale scintillation detectors for neutrino physics in general

and neutrinoless double beta decay in particular''

\notea{\bf Warms the cockles of our hearts...}\\

\noteq [Referee 2: General comments] `` ..however the discussion

should deserve...a deeper analysis.''

\notea{\bf Yes. We have extended the analysis to lower energies and

  have extracted much more detail from the simulation, and have

  extensively rewritten the manuscript. It is still at the

  `proof-of-principle level, as to go deeper yet will require vertex

  reconstruction, which is a much bigger effort (we are working on it).}\\

\noteq [Referee 2: General Comments] ``The physics case is included in

the title and mentioned in the text but never discussed in some

detail...A description of why the method is powerful for double beta

decay, and..to what extent it can improve the experimental sensitivity

would complete the discussion

\notea{\bf }\\

NYGREN'S RIGHT- CAN WE ADD A SENTENCE OR TWO TO THE INTRO AND

(PERHAPS) THE CONCLUSIONS?

\noteq [Referee 2: General Comments] ``The chosen example (5 MeV,

head-to-head single electrons) is of little interest for double beta decay.''

\notea{\bf We have added plots and numbers for 1.4 and 2.1 MeV, and

  substantially rewritten the text to differentiate between the cases

  of neutrino-electron scattering and double-beta decay.}\\

\noteq [Referee 2: General Comments] ``The dependence on the decay

position.. is not discussed.''

\notea{\bf To do a credible and useful job on the position dependence

  will require a fairly-detailed vertex reconstruction. This is beyond

  the scope of this present `proof-of-principle' paper. We have tried

  to clarify this in the text. }\\

\noteq [Referee 2: General Comments] Threshold effects as well as the

effect of statistical fluctuations for the low energy tails.. are not

discussed. The same holds for position and direction reconstruction

capabilities''

\notea{\bf }\\

I DON'T KNOW WHAT TO SAY HERE- SOMEBODY MORE EXPERT?

\noteq [Referee 2: Abstract] ``While appropriate for neutrino physics,

energy resolution is definitely worse with respect to other techniques..''

\notea{\bf We have fixed the text. Thanks.}\\

\noteq [Referee 2: Introduction] ``A reference ... would help to understand

the connection with Majorana neutrinos.''

\notea{\bf \D}\\

\noteq [Referee 2: Page 1] Is this 1 sigma or FWHM?

\notea{\bf It's sigma- we have fixed the text.}\\

\noteq [Referee 2: Page 2] ``better than water Cherenkov detectors''

\notea{\bf \FT}\\

PLEASE ADD `` for neutrino detectors'' after ``the other developed

technology'' AT THE TOP OF PAGE 2.

\noteq [Referee 2: Page 2 ] ``strong suppressant of backgrounds''..this

statement should deserve a deeper quantitative discussion ..''

\notea{\bf We agree. However, a useful ``quantitative discussion''

  will require a vertex reconstruction. We will get there, but at present

  it's beyond the scope of `proof-of-principle, looking to see if we

  can see the signal. We have tried to make this clearer in the text. }\\

\noteq [Referee 2: Equ. 2.1] ``..the meaning of the parameter `x' is

not described.''

\notea{\bf \F}\\

\noteq [Referee 2: Page 3] What is the typical cutoff value and the

corresponding useful fraction of Cherenkov light?

\notea{\bf We have added the cutoff values.}\\

DO WE HAVE THE USEFUL FRACTIONS OF CHERENKOV LIGHT? (I DON'T EVEN KNOW

IF THAT'S WELL-DEFINED?)

\noteq [Referee 2: Page 3] `` `Probable case being...' Events with

asymmetric electron energies are far more abundant.''

\notea{\bf \FT}\\

WE SHOULD REWRITE THIS SENTENCE ..E.G. ``We consider the case of

equal..'' OR SOMETHING.. (?) (MAYBE CHANGE 'PROBABLE' TO 'SIMPLE'?)

\noteq [Referee 2: Page 4] `` `Fully absorbing' ''. How are results

influenced by deviations from this hypothesis?''

\notea{\bf This is a level of detail well beyond what we've done so

  far; once we have a real detector design and a full simulation it

  can be addressed, but at this point devoting effort to it

doesn't seem productive.}\\

\noteq [Referee 2: Page 4] ``Single 5 MeV electrons''.

\notea{\bf \F}\\

\noteq [Referee 2: Page 4] ``I believe the `realistic' analysis

mentioned here is indispensable ..for this process''

\notea{\bf We agree, and are working on it. It isn't easy, or,

  unfortunately, fast compared to a post-doc tenure.}\\

\noteq [Referee 2: Page 6] ``The dependence on the position and the

corresponding transit time variations cannot in my opinion be neglected.''

\notea{\bf }\\

THIS NEEDS AN EXPERT'S ANSWER...

\noteq [Referee 2: Figure 3] `` This figure is very interesting

to understand the capacity of the method to maintain directionality

information... The comparison with a single event (or equivalently the

distribution of the variances) and the distribution for 1 MeV

electrons would improve to understand also the limits/power of the method''

\notea{\bf }\\

THIS NEEDS AN EXPERT'S ANSWER...

\noteq [Referee 2: reconstruction] ``It is not clear if limiting the

reconstruction to Cherenkov photons improve reconstruction''

\notea{\bf Yes, we agree. We are not proposing throwing away the

scintillation light, but in addition using the selected Cherenkov light to

determine the electron direction, i.e., not `limiting the

reconstruction to Cherenkov photons''}\\

CAN THIS ANSWER BE IMPROVED? (YES)

\end{document}

\noteq [Referee 1]

\notea{\bf }\\

### 2014-01-10 Direction_JINST_v5.tex
Added Christophs edits and committing Henrys referee response. --Lindley

\RequirePackage{lineno}

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the $\sim$1-2~MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
would have profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe\cite{leptogenesis}.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\% \sigma_{e}$/\sqrt{E(MeV)}$\cite{kam2013,borexino}. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, this energy resolution is much poorer than other technologies being used for $0\nu\beta\beta$ searches: Ge detectors\cite{gerda2013}, Te bolometers\cite{Alessandria:2011rc}, tracking detectors\cite{SuperNEMO}, liquid Xe time projection chambers (TPCs)\cite{EXO2012} and high pressure gaseous Xe TPCs\cite{NEXTsipm}.

Scintillation light is isotropic. At these low energies, it does not contain sufficient to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

This is the first in a series of papers exploring directionality in large-scale liquid scintillator detectors. In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application from the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.  We then use these results as input into a traditional direction reconstruction developed for water Cherenkov detectors. Since the reconstruction of the direction of $\sim$1~MeV particles has not been achieved before, we start these studies with the most simple case of a single particle at the center of the detector. We also start with an easier test case of a 5~MeV electron like that from a $^{8}$B solar neutrino interaction. With the higher photon statistics at this energy, we verify the technique and the different detector parameters that affect it. We then lower the energy to that relevant to a single electron in a $0\nu\beta\beta$.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{d^2N}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $N$ is the number of Cherenkov photons, $\lambda$ is the wavelength of the photon, x is the distance travelled by the particle, $Z$ is the charge of the particle, $\alpha$ is the fine structure constant, $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the particle.

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. For the example scintillator above, n(400nm)=1.459\cite{OlegThesis}. The absorption cutoff is between 370~nm and 360~nm where the attenuation length rapidly drops from 6.5~m to 0.65~m. For a 5~MeV electron, this yields 744 Cherenkov photons per event; dropping down to a 1~MeV electron yields 89 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm, the point above which a small number of photons is detected due to a standard photocathode's quantum efficiency. For reference at this cutoff wavelength, the index of refraction n(370nm)=1.466. This can be translated into an effective Cherenkov threshold for electrons at 0.188~MeV.

All photons including these undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
This means photons at the scintillation cutoff at 370~nm will travel at 0.191~m/ns while those at 600~nm travel at 0.203 m/ns. Since on average the undisturbed Cherenkov photons have longer wavelengths, they will arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying $0\nu\beta\beta$ decay mechanism\cite{SuperNEMO,newphysics0nuBB,bandv}.
For light Majorana neutrino exchange, a probable case is the equal division of energy between two
electrons emitted at a large angle relative to each other\cite{phasespace, SuperNEMO}. An example case is shown in figure \ref{detector_view}
for a $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred. Due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains, isotopes with Q-values above the 2.6~MeV gamma from $^{208}$Tl are preferred; however slightly lower Q-values can be tolerated if there is sufficient background rejection. There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with sufficiently large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently nanocrystals formed by candidate isotopes have been explored as an alternative to doping by single atoms,\cite{qdot,qdot2}.

@@

Re-emission of absorbed photons in the scintillator
bulk volume and optical scattering, specifically Rayleigh scattering, have not yet been included by default. A test simulation shows that the effect of optical scattering is negligible. The reason is that the cutoff is very steep below 360~nm and almost no photons reach the sphere, so optical scattering makes no difference for short wavelengths. Above about 395~nm, the attenuation length is greater than 20~m so both scattering and absorption are not very likely and scattering is negligible. The intermediate region is rather small. A similar argument holds for re-emission. Scattering length measurements and discussions can be found in Ref.~\cite{Wurm:2010ad}.

@@

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut, 171~PEs
        from scintillation and 108~PEs from Cherenkov light are detected. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut, 349~PEs from
        scintillation and 88~PEs from Cherenkov light are detected. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut, 226~PEs from scintillation and 229~EPS
        from Cherenkov light are detected. \label{time_plots_comparison}}
        \end{center}
\end{figure}

>>>>

%\RequirePackage{lineno}

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
would have profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe\cite{leptogenesis}.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%$/\sqrt{E(MeV)}$\cite{kam2013,borexino}. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, this energy resolution is much poorer than other technologies being used for $0\nu\beta\beta$ searches: Ge detectors\cite{gerda2013}, Te bolometers, liquid Xe time projection chambers (TPCs)\cite{EXO2012} and high pressure gaseous Xe TPCs\cite{NEXTsipm}.

Scintillation light is isotropic and any information on the primary particle's direction is lost.  Cherenkov light is also
produced. Most is absorbed and re-emitted as part of the
scintillation processes, however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

This is the first in a series of papers exploring directionality in large-scale liquid scintillator detectors. In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector advances in timing, photodetector 
spectral response, and scintillator emission spectra that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors. This is a different technique and application than the direction reconstruction described for high-energy neutrino interactions\cite{john} or that for neutrons from inverse beta decay\cite{chooz,dcDirection}.  We then use these results as input into a traditional direction reconstruction developed for water Cherenkov detectors. Since the reconstruction of the direction of $\sim$1~MeV particles has not been achieved before, we will start these studies with the most simple case of a single particle at the center of the detector. We will also start with an easier test case of a 5~MeV electron like that from a $^{8}$B solar neutrino interaction. With the higher photon statistics at this energy, we will verify the technique and the different detector parameters that effect it. We will then lower the energy to that relevant to a single electron in a $0\nu\beta\beta$.  Since these simpler scenarios are successful, this justifies future work, specifically on the reconstruction of the two simultaneous electrons throughout the detector volume and the study of the sensitivity of a detector instrumented to take advantage a directional signal to both $0\nu\beta\beta$ and to new physics.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{d^2N}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where x is the distance travelled by the particle, $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle.

The Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. For the example scintillator above, n(400nm) is 1.459\cite{OlegThesis}. The absorption cutoff is between 360~nm and 370~nm where the attenuation length rapidly drops from 6.5~m to 0.65~m. For a 5~MeV electron, this yields 744 Cherenkov photons per event, and dropping down to a 1~MeV electron yields 89.3 Cherenkov photons per event when integrating from the cutoff wavelength at 360~nm to 550~nm, the point above which a negligible number of photons is produced.

All photons including these undisturbed Cherenkov
photons will have timing determined by the group 
velocity\cite{group_velocity_article,pdg_review_2012,tamm1939} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
This means photons right above the scintillation cutoff at 370~nm will travel at 0.1914~m/ns while those at longer wavelengths for example 600~nm travel at 0.2031 m/ns. Since on average the undisturbed Cherenkov photons have longer wavelengths, they will arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons follow distributions of energies and angular correlations which depend on the underlying decay $0\nu\beta\beta$ mechanism\cite{newphysics0nuBB, SuperNEMO}.
For the basic mechanism of light Majorana neutrino exchange, a probable case is the equal division of energy between back-to-back
electrons\cite{phasespace, SuperNEMO}. This case is shown in figure \ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred. Due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains, isotopes with Q-values above the $^{208}$Tl  2.6~MeV gamma are preferred, however slightly lower Q-values can be tolerated if there is sufficient background rejection. There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with sufficiently large Q-values.

Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd (Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr (Q=3.350~MeV)\cite{zr1},
$^{100}$Mo (Q=3.034~MeV)\cite{mo1}, $^{82}$Se (Q=2.995~MeV)\cite{qdot},
$^{116}$Cd (Q=2.81~MeV)\cite{qdot, cd1}, $^{130}$Te (Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe (Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn (Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
a suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. As an alternative to single-atom-doping, recently nanocrystals 
formed by candidate isotopes have been explored as a dopant\cite{qdot,qdot2}.

@@

Re-emission of absorbed photons in the scintillator
bulk volume and optical scattering, specifically Rayleigh scattering, have not yet been included by default. A test simulation shows that their effects are negligible. The reason is that the cutoff is very steep below 360~nm and almost no photons reach the sphere, so optical scattering makes no difference. Above about 395~nm, the attenuation length is greater than 20~m so both scattering and absorption are not very likely and scattering is negligible. The intermediate region is rather small.  This is the conclusion of a more detailed study found in Ref.~\cite{Wurm:2010ad}.

@@

\begin{figure}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cherenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dash-dotted vertical line illustrates a time cut at
        34.0~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cherenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cherenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cherenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure}
